Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-04-25 manuscript/manuscript.tex
lots of edits to text, plots

\begin{abstract}
Supermassive black hole binaries (SMBHBs) are predicted to form a 
high-redshift cosmological population of low-frequency gravitational wave (GW) sources, 
which added together produce a stochastic gravitational wave background (GWB). 
Pulsar timing arrays (PTAs) can observe this signal by looking for correlations in the 
times of arrival (TOAs) of millisecond pulsars (MSPs). 
In this paper, we introduce an improved frequentist estimator called the noise-marginalized optimal statistic, 
which can be used to assess the significance of a GWB present in PTA data. 
We find \ki{complete}

\end{abstract}

@@

Long-wavelength gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \cite{hd1983,fb1990}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival. 
Current upper-limits on the stochastic background from PTAs are approaching 
theoretical predictions for the stochastic GWB \citep{PPTA2013,EPTA2015,NANOGrav9yr}

@@

However, previous work has found that the optimal statistic gives biased results 
for PTAs with significant red noise in the individual pulsars due to the 
strong covariance between the individual red noise parameters 
and the GWB amplitude \sv{has this ever been discussed in a paper?}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: marginalizing over the individual pulsar red noise parameters using the posterior distributions from a full PTA Bayesian analysis. This hybrid approach produces a more precise estimate of the stochastic GWB amplitude, 
\Agw\/, and its uncertainty, while requiring only a few minutes over the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization can be used to compute the optimal statistic with any choice of overlap-reduction function (ORF), allowing for quick testing of various spatial correlations.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic 
and compare to the fixed-noise optimal statistic. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between different spatial correlations 
using simulated PTA data. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
%We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We conclude in Sec.~\ref{sec:conclusion}.

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details). 
To these data sets we injected an isotropic stochastic GWB 
\sv{insert strength of GWB... are we going to have one value or more than one?}.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs.}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

The optimal statistic was computed using the PTA data analysis package 
PAL2 %\footnote{\href{https://github.com/jellis18/PAL2}{https://github.com/jellis18/PAL2}} 
\citep{evh17a}. 
Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
%	\caption{Optimal statistic and SNR for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters (solid orange line). 
%			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%			for a common red process (solid green line). 
%			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			For this particular realization of the stochastic background, the fixed-noise analysis 
%			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
%			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
%			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
%	\label{fig:os_dataset_sample}
%\end{figure}

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
%	\caption{Optimal statistic and SNR for 425 simulated datasets 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (blue) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters. The dashed vertical line 
%			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			The value of $\hat{A}^2$ from the fixed-noise analysis, 
%			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
%			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
%			noise-marginalized analysis gives an average mean SNR of 1.7.}
%	\label{fig:os_datasetstats}
%\end{figure}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in $\tilde{{\bf{S}}}_{IJ}$. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
$\Gamma(\theta_{IJ}) = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma(\theta_{IJ}) = \cos\theta_{IJ}$.

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORFs, respectively.
%\begin{figure}[htb]
%	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
%	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
%			containing an injected isotropic stochastic GWB with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using monopole (blue), dipole (orange), 
%			and quadrupole (green) spatial correlations. 
%			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
%	\label{fig:os_ORF}
%\end{figure}

@@

\acknowledgments
We thank Joe Romano and Xavier Siemens for useful discussions. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA.

>>>>

\begin{abstract}
Observations have revealed that nearly all galaxies contain supermassive black holes (SMBHs) at their centers, 
with masses between \sv{look up}. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these GW sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator for the amplitude of a GWB 
that specifically looks for the spatial correlations induced by the GWB in PTA data. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of a GWB, 
and we use the noise-marginalized optimal statistic to assess the 
significance of different spatial correlations and 
by performing sky scrambles.
\end{abstract}

@@

Long-wavelength gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \cite{hd1983,fb1990}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival. 
Current upper-limits on the stochastic background from PTAs are approaching 
theoretical predictions for the stochastic GWB \citep{PPTA2013,EPTA2015,abb+17b}

@@

However, the optimal statistic gives biased results 
for PTAs with significant red noise in the individual pulsars due to the 
strong covariance between the individual red noise parameters 
and the GWB amplitude \sv{has this ever been discussed in a paper?}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the stochastic GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes over the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic with any choice of overlap-reduction function (ORF), 
allowing for quick testing of various spatial correlations. 
This technique was used to perform the frequentist searches 
for common red signals with monopole, dipole, and Hellings-Downs spatial correlations 
in the NANOGrav 11-year dataset \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed using fixed noise. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
%We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
and discuss future applications of the noise-marginalized optimal statistic.

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details). 
To these data sets we injected an isotropic stochastic GWB 
\sv{insert strength of GWB... are we going to have one value or more than one?}.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs. \sv{finish filling in this table}}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

The optimal statistic was computed using the PTA data analysis package 
PAL2 %\footnote{\href{https://github.com/jellis18/PAL2}{https://github.com/jellis18/PAL2}} 
\citep{evh17a}. 
Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background with $\Agw=$. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
%	\caption{Optimal statistic and SNR for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters (solid orange line). 
%			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%			for a common red process (solid green line). 
%			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			For this particular realization of the stochastic background, the fixed-noise analysis 
%			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
%			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
%			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
%	\label{fig:os_dataset_sample}
%\end{figure}

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis more accurately recovers the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis using the individual noise results, 
averaged over realizations of the GWB, is $\hat{A}^2 = 7.9 \times10^{-30}$ with a mean SNR of 2.3. 
The fixed-noise and noise-marginalized analyses using the common noise results both give 
$\hat{A}^2 = 2.4\times10^{-29}$ and a mean SNR of 4.1.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated datasets 
			containing an injected GWB. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			We compare the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			to the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis more accurately estimates $\hat{A}^2$.}
	\label{fig:os_datasetstats}
\end{figure}

The fixed-noise and noise-marginalized analyses gave the same results 
for an injected GWB with $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
Figure~\ref{fig:os_datasetstats2} shows the mean optimal statistic and mean SNR 
for 100 realizations of a GWB with $\Agw = 10^{-15}$. 
The fixed-noise analysis using the individual noise values 
gives a mean value of $\hat{A}^2 = 2.9\times10^{-31}$ and a mean SNR of 1.1. 
The fixed-noise analysis using the common noise values 
gives a mean value of $\hat{A}^2 = 8.4\times10^{-31}$ and a mean SNR of 2.1, 
which the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 1.1\times10^{-30}$ and a mean SNR of 2.4.  
The difference between the fixed-noise and noise-marginalized analyses using the common noise values 
is very small compared to the large spread in $\hat{A}^2$ and the SNR over different realizations of the GWB.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A1e-15.pdf}
	\caption{Optimal statistic and SNR for 100 simulated datasets 
			containing an injected GWB. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 10^{-30}$. 
			We compare the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			to the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis more accurately estimates $\hat{A}^2$.}
	\label{fig:os_datasetstats2}
\end{figure}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in Eq.~\eqref{eq:}. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
$\Gamma(\theta_{IJ}) = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma(\theta_{IJ}) = \cos\theta_{IJ}$.

Our ability to distinguish between different spatial correlations 
depends on the strength of the GWB 
and the angular separations between pulsar pairs, $\theta_{IJ}$. 
We can determine the overlap between ORFs corresponding to different spatial correlations 
by computing the ``match statistic'' \citep{cs2016},
\begin{equation}
	\bar{M} = \frac{\sum_{I,J \neq I} \Gamma_{IJ} \Gamma'_{IJ}}{\sqrt{ \left( \sum_{I, J \neq I} \Gamma_{IJ} \Gamma_{IJ} \right) \left( \sum_{I, J \neq I} \Gamma'_{IJ} \Gamma'_{IJ} \right)}} \,,
\end{equation}
where $I$,$J$ index the pulsar number, and $\Gamma$ and $\Gamma'$ are two different ORFs. 
For the 18 pulsars used in these simulations, the 
match statistic for monopole and Hellings-Downs correlations is $\bar{M} = 0.264$, 
and the match statistic for dipole and Hellings-Downs correlations is $\bar{M} = 0.337$. 
These match statistics describe a fundamental limit on our ability 
to identify the spatial correlations of a common red signal as Hellings-Downs 
rather than monopole or dipole 
that depends only on the number of pulsars in our PTA and their sky positions.

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 200 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the quadrupole ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the quadrupole ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[htb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 200 simulated datasets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

\acknowledgments
We thank Joe Romano, Xavier Siemens, and Jeff Hazboun for useful discussions. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA.
